---
title: "Cervical Cancer Classification and Handling Class Imbalance"
author: "Yupeng He & Martin Li"
date: "2024-09-26"
output:
  pdf_document:
    toc: true
    toc_depth: 2
  html_document:
    toc: true
    toc_depth: 2
    highlight: "zenburn"
---

# Introduction

This project focuses on classifying cervical cancer using a dataset of risk factors and applying various machine learning techniques to handle class imbalance.

## Dataset Overview

The dataset includes risk factors for cervical cancer, and the goal is to classify the presence or absence of cancer (Dx) using features like age, STDs, and biopsy results.

```{r setup, include=FALSE}
# Load necessary libraries
library(tidyverse)
library(caret)
library(randomForest)
library(gbm)
library(UBL)
library(pROC)
```

### Load Data

We start by loading the dataset and replacing missing values with the median for numerical features.

```{r}
# Load the dataset
data <- read.csv('data/risk_factors_cervical_cancer.csv')

# Replace '?' with NA and handle missing values
data[data == "?"] <- NA
data <- as.data.frame(sapply(data, as.numeric))
data$Dx <- as.factor(data$Dx)

# Replace NAs with median values for numeric columns
data <- data %>%
  mutate(across(where(is.numeric), ~ coalesce(., median(., na.rm = TRUE))))

# Remove columns that are not relevant
data <- data %>% select(-Dx.Cancer)
data$Dx <- as.factor(data$Dx)
levels(data$Dx) <- c("No Cancer", "Cancer")
```

### Exploratory Data Analysis

We visualize the distribution of cancer diagnoses in the dataset and check the top 8 correlations with the target variable.

```{r}
# Plot the distribution of cancer diagnosis
ggplot(data, aes(x = Dx)) +
  geom_bar(fill = "steelblue") +
  theme_minimal() +
  ggtitle("Distribution of Cancer Diagnosis")

# Correlation plot for top correlated variables
cor_matrix <- cor(data %>% select(-Dx))
top_cor <- names(sort(abs(cor_matrix["Dx",]), decreasing = TRUE)[1:9])
ggplot(melt(cor_matrix[top_cor, top_cor]), aes(X1, X2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), color = "white", size = 3) +
  scale_fill_gradient2(low = "blue", high = "red") +
  ggtitle("Top Correlated Features with Dx")
```

## Data Preprocessing

We split the data into training and testing sets and scale numeric features.

```{r}
# Stratify sampling to handle class imbalance
set.seed(1)
train_index <- createDataPartition(data$Dx, p = 0.7, list = FALSE)
train <- data[train_index, ]
test <- data[-train_index, ]

# Scale numeric columns
train <- train %>% mutate(across(where(is.numeric), scale))
test <- test %>% mutate(across(where(is.numeric), scale))
```

## Model Training

We train multiple machine learning models to classify cervical cancer, including logistic regression, k-NN, bagging, random forest, and boosting. We evaluate model performance using cross-validation.

```{r}
# Define a function to perform 5x5 cross-validation
cv_accuracy <- function(model, train, k = 5) {
  set.seed(1)
  cv_index <- createMultiFolds(train$Dx, k = k, times = k)
  results <- sapply(names(cv_index), function(fold) {
    fold_train <- train[cv_index[[fold]], ]
    fold_test <- train[-cv_index[[fold]], ]
    pred <- model(fold_train, fold_test)
    confusionMatrix(as.factor(pred), fold_test$Dx)$overall['Accuracy']
  })
  mean(results)
}

# Logistic regression model
logistic <- function(train, test) {
  model <- glm(Dx ~ ., data = train, family = "binomial")
  pred <- predict(model, newdata = test, type = "response")
  ifelse(pred > 0.5, "Cancer", "No Cancer")
}

# Random forest model
random_forest <- function(train, test) {
  model <- randomForest(Dx ~ ., data = train)
  predict(model, newdata = test)
}

# Boosting model
boosting <- function(train, test) {
  model <- gbm(Dx ~ ., data = train, distribution = "multinomial", n.trees = 5000)
  pred <- predict(model, newdata = test, n.trees = which.min(model$cv.error), type = "response")
  ifelse(pred > 0.5, "Cancer", "No Cancer")
}

# Cross-validation results
results <- data.frame(
  Model = c("Logistic", "Random Forest", "Boosting"),
  Accuracy = c(cv_accuracy(logistic, train), cv_accuracy(random_forest, train), cv_accuracy(boosting, train))
)
results
```
## Addressing Class Imbalance with ADASYN

We handle the class imbalance by applying the ADASYN algorithm for synthetic sampling.

```{r}
# Apply ADASYN for oversampling
train_balanced <- AdasynClassif(Dx ~ ., train, beta = 1, k = 5)

# Check the distribution after oversampling
ggplot(train_balanced, aes(x = Dx)) +
  geom_bar(fill = "steelblue") +
  theme_minimal() +
  ggtitle("Distribution After ADASYN Oversampling")
```

## Model Evaluation

We evaluate model performance using accuracy, precision, recall, and F1-score. ROC curves are plotted for a final comparison of the models.

```{r}
# Evaluate the models on the test set
predictions <- list(
  Logistic = logistic(train, test),
  RandomForest = random_forest(train, test),
  Boosting = boosting(train, test)
)

# Calculate confusion matrices
conf_matrices <- lapply(predictions, function(pred) {
  confusionMatrix(as.factor(pred), test$Dx)
})

# ROC curves
roc_curves <- lapply(predictions, function(pred) {
  roc(test$Dx, as.numeric(pred == "Cancer"))
})

# Plot ROC curves
ggroc(roc_curves, aes(color = names(predictions))) +
  ggtitle("ROC Curves for Different Models")
```

# Conclusion

The application of various machine learning models to the cervical cancer dataset shows that random forest and boosting models perform the best. ADASYN helped in mitigating the class imbalance issue, improving the models’ ability to predict cervical cancer cases.